\section{Not included: Tree-like Queries}

\diego{Tree patterns as CRPQs: define "standard encoding" of letters by self-loops}
The minimization of ""tree patterns"" has been thoroughly studied \cite{min-tree-patterns}. 
A "tree pattern" can be seen in our context as a special kind of tree-like "CRPQ". Let us call a \AP""pseudo-tree"" to any directed multigraph such that (i) it has no parallel edges (in particular no parallel self-loops) and (ii) once the self-loops are removed, it is a directed `root-to-leaves' tree.
Let us fix an alphabet letter $e \in \A$ which will be used to model the child relation.
A \AP""tree pattern CRPQ"" is then any "CRPQ" such that
\begin{itemize}
  \item the underlying directed multigraph of the query is a "pseudo-tree"; and
  \item its "self-loops" are of the form $x \atom{a} x$ for some $a \in \A \setminus \set{e}$
  \item its remaining "atoms" are form (a) $x \atom{e^+} y$ or (b) $x \atom{e} y$.
\end{itemize}
A \AP""pseudo-tree database"" is a "graph database" such that
\begin{itemize}
  \item the underlying directed multigraph of the query is a "pseudo-tree"; and
  \item every "vertex" $v$ there is an edge $(v,a,v)$ for some $a \in \A \setminus \set{e}$
  \item its remaining "edges" are of the form $(v,e,v')$.
\end{itemize}

\knowledgenewrobustcmd{\treePatEnc}[1]{\cmdkl{\textup{\bf t}_{#1}}}
We can then define a bijection between "tree pattern CRPQs" and "tree patterns". Given a "tree pattern CRPQ" $q$, we define $\intro*\treePatEnc{q}$ to be the "tree pattern" consisting of treating self-loops as node labels, $x \atom{e^+} y$ as the descendant relation, and $x \atom{e} y$ as the child relation. It is easy to see that $\set{q \mapsto \treePatEnc{q}}$ is a bijection. 

\begin{lemma}\AP\label{lem:treepat-crpq}
  A "tree pattern CRPQ" $q$ is "atom-minimal" if, and only if, $\treePatEnc{q}$ is minimal (among tree patterns over trees).
\end{lemma}
\begin{proof}
  \proofcase{Left-to-right.}
  Observe the following:
  \begin{claim}\AP\label{cl:atommin-over-trees=treepatmin}
    The following are equivalent for each "tree pattern CRPQ" $q$:
    \begin{enumerate}
      \item $q$ is "atom-minimal" among "tree pattern CRPQs" over "pseudo-tree databases";
      \item $\treePatEnc{q}$ is minimal among "tree patterns" over trees.
    \end{enumerate}  
  \end{claim}
  \begin{claim}\AP\label{cl:equiv-overtrees-overgraphs}
    Two "tree pattern CRPQs" are "equivalent" over "pseudo-tree databases" if, and only if, they are "equivalent" over arbitrary "graph databases".
  \end{claim}
  In view of \Cref{cl:atommin-over-trees=treepatmin,cl:equiv-overtrees-overgraphs} we have that
  % \begin{remark}
    if a "tree pattern CRPQ" $q$ is "atom-minimal" (among all "CRPQs" over arbitrary "graph databases"), then $\treePatEnc{q}$ is "minimal" (among "tree patterns" over trees).
  % \end{remark}

  \proofcase{Right-to-left.}
  By \Cref{cl:atommin-over-trees=treepatmin,cl:equiv-overtrees-overgraphs}, it suffices to show that if a "tree pattern CRPQ" is "atom-minimal" among "tree pattern CRPQs", then it is also "atom-minimal" among all "CRPQs". In other words, if a "tree pattern CRPQs" $q$ is "equivalent" to a "CRPQ" $q'$ with fewer "atoms", then it is also equivalent to a "tree pattern CRPQ" $q''$ with fewer "atoms".
  \diego{TBC... Rémi, the floor is yours :)}
\end{proof}

As a consequence of \cite[Theorem~3.1]{min-tree-patterns} and the previous \Cref{lem:treepat-crpq} we then have:
\begin{corollary}
  The "atom-minimization problem" for "tree pattern CRPQs" is "SigmaP2"-complete.
\end{corollary}