
% TREE PATTERNS (SUB?)SECTION HERE:
\input{sec-treepatterns}  
\section{other things removed}
\paragraph{Strong Redundancy}
\diego{Remove an "atom" $\alpha = x \atom{L} y$ of a "CRPQ" $q$ if $L$ contains the language induced by the concatenation of the languages on a path from $x$ to $y$ on $q \setminus \set{\alpha}$. We obtain $q \semequiv q \setminus \set\alpha$. It is therefore in "PSpace" to test if there is a "strongly redundant" "atom".}
\paragraph{Completion}
\diego{adding an atom between two existing variables with the induced langauge. Granted, this actually increases the number of atoms! But in conjunction with "redundancy" or even "weak-redundancy" it may lead to smaller queries. In the example of \Cref{fig:ex-equiv-queries} we can see the passage from left to right as the result of performing: first completion, then redundancy elimination and last contraction.}
\paragraph{Weak Redundancy}
\diego{We can apply Completion, Contraction and (strong) Redundancy. Is there a limit on how many times these have to be applied?}

A ""weak-homomorphism"" from a "CRPQ" $q$ to another "CRPQ" $q'$ is a mapping $h: \vars(q) \to \vars(q')$ such that for every atom $x \atom{L} y$ of $q$ there is a path $h(x) \atom{L_1} \dotsb \atom{L_n} h(y)$ from $h(x)$ to $h(y)$  in $q'$ where $L_1 \dotsb L_n \subseteq L$.
\begin{lemma}
  The "weak-homomorphism" relation is reflexive and transitive.
  If there is a "weak-homomorphism" from $q$ to $q'$, then $q' \contained q$.
\end{lemma}
An "atom" $\alpha$ of $q$ is ""weakly-redundant"" if there is a "weak-homomorphism" from $q$ to $q \setminus \set\alpha$.
\diego{Question: is there a notion of ``core''?}

=====================

Say that a class $\classCRPQ$ of CRPQs satisfies the \AP""uniform small model property""
if for every $\gamma \in \classCRPQ$, there exists a function
$\intro*\shrink{\gamma}\colon \Exp(\gamma) \times \N \to \Exp(\gamma)$ "st":
\begin{enumerate}
  \item for each $\gamma \in \classCRPQ$, for each $n \in \N$ and $\delta \in \classCRPQ$,
  if $\nbatoms{\delta} \leq n$ then for any $G \in \Exp(\gamma)$, $G \FOmodels \delta$ "iff"
  $\shrink{\gamma}(G, n) \FOmodels \delta$, and
  \item the size of $\shrink{\gamma}(G, n)$ only depends on $\gamma$ and $n$---but not $G$.
\end{enumerate} 
Note that $\shrink{\gamma}(G, n) \defeq G$ trivially satisfies the first point but not the second one.
In the name "uniform small model property" ``uniform'' refers to the fact
that the way we shrink $G$ onto $\shrink{\gamma}(G,n)$ is the same for all $\delta$---provided that $\nbatoms{\delta} \leq n$. The ``small model property'' part refers to the fact that to check if
$\gamma \contained \delta$, it suffices to check that all canonical databases of $\gamma$
of size at most $\nbatoms{\shrink{\gamma}(-, \nbatoms{\delta})}$ are models of $\delta$.

\begin{proposition}
  \AP\label{prop:containment-uniform-smp}
  If $\classCRPQ$ has the "uniform small model property", if $\gamma \in \classCRPQ$ and $\Delta$ 
  is a finite union of $\classCRPQ$-queries, then
  $\gamma \contained \Delta$ "iff" for every "canonical database" $G$ of $\gamma$,
  $\shrink{\gamma}(G, m) \FOmodels \Delta$
  where $m \defeq \max{\set{\nbatoms{\delta} \mid \delta \in \Delta}}$.
\end{proposition}

\begin{proof}
  The left-to-right implication is trivial. For the right-to-left implication, 
  let $G$ by any "canonical database" of $\gamma$,
  and assume by contradiction that $G \not\FOmodels \Delta$.
  Then for all $\delta \in \Delta$, $G \not\FOmodels \delta$,
  and so $\shrink{\gamma}(G, m) \not\FOmodels \delta$.
  It follows that $\shrink{\gamma}(G, m) \not\FOmodels \Delta$.
  Contradiction.
\end{proof}

\begin{proposition}
  \AP\label{prop:tree-patterns-uniform-smp}
  The class of tree patterns---and more generally CRPQs over $(a, a^+)$ or even over $(a, a^*)$---
  has the "uniform small model property", where $\shrink{\gamma}(G, n)$ consists
  in shrinking every "atom refinement" of size at least $n+1$ to an "atom refinement" of
  size $n$.
\end{proposition}

\begin{proof}
  TODO. Addref?
\end{proof}

\begin{corollary}
  \AP\label{coro:containment-union-tree-patterns-better}
  The containment problem of deciding if $\gamma \contained \Delta$, given:
  \begin{itemize}
    \item an integer $n$,
    \item a "tree pattern" $\gamma$ of size at most $n$,
    \item a union of "tree patterns" $\Delta$ "st" each $\delta \in \Delta$ has size at most $n$
    and deciding if a "tree pattern" $\delta$ belongs to $\Delta$ can be done in "NP" ("wrt" $n$),
  \end{itemize} 
  can be solved in "PiP2".
\end{corollary}

\begin{proof}
  By \Cref{prop:containment-uniform-smp,prop:tree-patterns-uniform-smp},
  $\gamma \contained \Delta$ "iff" for every "canonical database" $G$ of $\gamma$,
  there exists $\delta \in \Delta$ "st"
  $\shrink{\gamma}(G, n) \FOmodels \delta$.
  In turn, this is equivalent to asking that for every "canonical database" $G$ of $\gamma$
  of size at most $|\shrink{\gamma}(-, n)|$, then there exists $\delta\in \Delta$
  "st" $G \FOmodels \delta$.
  This concludes the proof that the problem lies in "PiP2" since by \Cref{prop:tree-patterns-uniform-smp}, $\nbatoms{\shrink{\gamma}(-, n)} = n^2$ so this
  upper bound is polynomial.
\end{proof}

\begin{proof}[Proof of \Cref{proof:union-tree-patterns}]
  \proofcase{Upper bound.}
  A "tree pattern" $\gamma$ is equivalent to a finite union of smaller "tree patterns"
  "iff" $\gamma \contained \App{\gamma}{\+C}{\textit{poly}}$ by \Cref{prop:max-under-approx},
  where $\+C$ is the class of all trees which are strictly smaller as "tree patterns"
  than $\gamma$.
  \remi{TODO: well not exactly…
  Also: size of tree pattern $\neq$ size of its CRPQ encoding (because of node labels). Solution: consider classes of queries, not of graphs.}
  But then \remi{also argue this point}
  \[
    \App{\gamma}{\+C}{\textit{poly}} \semequiv 
    \{
      \alpha \in \CRPQ[\+C] \mid \exists \rho \in \Refin[\leq m](\gamma), 
      \rho \surjto \alpha
    \}.
  \]
  Each $\alpha$ has polynomial (in fact linear) size,
  and testing if $\alpha$ belongs to this latter union can be done in "NP" by definition,
  so \Cref{coro:containment-union-tree-patterns-better} provides an algorithm in "PiP2"
  for the problem.

  \proofcase{Lower bound.} \remi{todo}
\end{proof}

\remi{Check if it also works for "CRPQs" over simple regular expressions.}

\remi{What about minimization of variables?}
\input{sec-treelike}

\section{Removed from Discussion}
\subsection{Tree patterns} Tree pattern minimization on the class of tree patterns is "PiP2". Tree pattern minimization on the class of unions of tree patterns may be better. Possibly the example of \cite[Fig.~9]{CzerwinskiMartensNiewerthParys2018Minimization} is minimal, and yet it is equivalent to a union of small conjunctive queries.
Our results on "SRE@@positive" yield that unions of tree pattern minimization is still "PiP2", and hence that by going to a more expressive formalism we don't lose in complexity.

\begin{conjecture}
If a tree-like CRPQ is equivalent to a CRPQ of size $k$, then it is equivalent to a tree-like CRPQ of size $k$.
\end{conjecture}



\begin{openproblem}
    What's the exact complexity of CRPQ (atom or variable) minimization?
\end{openproblem}

\paragraph{Graphs with transitive edges} A very simple query language is the case of graphs with edges and transitive edges, which naturally corresponds to CRPQs having only two possible languages $a$ and $a^+$. In this simple case the upper bound for CRPQ minimality can be improved to XXX.

\begin{openproblem}
    Does minimization of CRPQ on SRE have a better complexity than "2ExpSpace"?
\end{openproblem}