\chapter{The Algebras for Automatic Relations}
\AP\label{ch:algebra}
\renewcommand\thefigure{\thechapter.\arabic{figure}}

\begin{chapterpresentation}
	\begin{abstract}
		We introduce ``synchronous algebras'', an algebraic structure tailored to recognize automatic relations ("aka" automatic relations, or regular relations). They are the equivalent of monoids for regular languages, however they conceptually differ in two points: first, they are typed and second, they are equipped with a dependency relation expressing constraints between elements of different types.

		The interest of the proposed definition is that it allows to lift, in an effective way, pseudovarieties of regular  languages to that of automatic relations, and we show how algebraic characterizations of pseudovarieties of regular languages can be lifted to the pseudovarieties of automatic relations that they induce.
		Since this construction is effective, this implies that the membership problem is decidable
		for (infinitely) many natural classes of automatic relations.
		A typical example of such a pseudovariety is the class of ``group relations'', defined as the relations recognized by finite-state synchronous permutation automata.
	
		In order to prove this result, we adapt two pillars of algebraic language theory to synchronous algebras: (a) any relation admits a syntactic synchronous algebra recognizing it, and moreover, the relation is synchronous if, and only if, its syntactic algebra is finite and (b) classes of automatic relations with desirable closure properties ("ie" pseudovarieties) correspond to pseudovarieties of synchronous algebras.
	\end{abstract}
	\par\bigskip\bigskip
	\begin{acknowledgements}
		This article is mostly a reproduction of \cite{Morvan2025Algebras}, which was published at
		CSL '25, with minor modifications to the introduction and conclusion, and
		appendices have been incorporated to the main body.
		\Cref{sec:theorem-projection-monad-adjunction} is entirely new.
		We thank Pablo Barceló, Mikołaj Bojańczyk, and Diego Figueira for
		helpful discussions, and some anonymous reviewers for valuable feedback.
	\end{acknowledgements}
	\clearpagepresentation
	\chaptertocstandalone
\end{chapterpresentation}

\input{parts/algebra/sec-introduction}
\input{parts/algebra/sec-preliminaries}
\input{parts/algebra/sec-synchronous-algebras}
\input{parts/algebra/sec-lifting-theorem}
\input{parts/algebra/sec-discussion}

\clearpage
\begin{subappendices}
	\input{parts/algebra/apdx-monads}
	% \input{parts/algebra/apdx-pointers}
\end{subappendices}